A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-01-22 ManuscriptGeophysics2018.tex
Version of initias submission to Geophysics

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
%
% (1) PRINCIPAL OBJECTIVES AND SCOPE OF THE WORK
  The open-source code \texttt{fdesign} makes it possible to design digital
  linear filters for the Hankel and Fourier transforms used in potential-,
  diffusive-, and wavefield modeling. Digital filters can be derived for any
  electromagnetic method, such as methods in the diffusive limits (DC, CSEM) as
  well as methods using higher frequency content (GPR, acoustic and elastic
  wavefields).
%
% (2) METHODOLOGY
  The direct matrix inversion method is used for the derivation of the filter
  values, and a brute-force inversion is carried out over the defined spacing
  and shifting values of the filter basis. Included or user-provided
  theoretical transform pairs are used for the inversion. Alternatively, one
  can provide 1D subsurface models that will be computed with a precise
  quadrature method using the EM modeler \texttt{empymod} to generate
  numerical transform pairs.
%
% (3) RESULTS
  The comparison of the presented 201\,pt filter with previously presented
  filters shows that it performs better for some standard CSEM cases. The
  derivation of a longer 2001\,pt filter for a GPR example with 250\,MHz center
  frequency proves that the filter method works also for wave phenomena, not
  only for diffusive EM fields.
%
% (4) CONCLUSIONS
  The presented algorithm provides a tool to create problem specific digital
  filters. Such purpose-built filters can be made shorter and can speed up
  consecutive  potential-, diffusive-, and wavefield inversions.
%
\end{abstract}

@@

\subsection{Minimization criteria}
Figure~\ref{fig:InvCrit} shows the differences between the different
minimization approaches; in (a) for a conventional, fast decaying transform
pair; and in (b) for a more complex, high frequency layered model. The circles
show the minimum amplitude used in previous approaches. This criteria is not
ideal, as it is subject to some random fluctuations and also depends on the
choice of $r$. The squares show the minimum amplitude given a certain
acceptable error, and the diamonds show the maximum $r$ given a certain
acceptable error (the inversion is a minimization process, it therefore
minimizes $1/r$, not $r$). In simple cases the minimum amplitude given an
acceptable error and the maximum $r$ given an acceptable error will yield the
same result. However, in complex cases the maximum $r$ is more consistent and
therefore a much better criteria.
%
\plot*{InvCrit}{width=.9\textwidth}{(a) A regular RHS curve of a transform pair
  with a purely decaying function. Minimum amplitude or maximum $r$ yield the
  same result in this scenario. Using a relative error criteria is more stable
  than just the absolute minimum amplitude. (b) A 1D model for $f=200\,$MHz.
  Here, the maximum $r$ provides a better criteria for the inversion.}
%

@@

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often  a good filter is found by shear luck of
choosing the right starting parameters. Each of the input parameters has an
effect to the outcome. It also depends heavily on the transform pairs
$f_\mr{I;C}$; functions that decay rapidly are generally better, as noted by
earlier authors \citep[e.g.][]{USGS.75.Anderson}. It also depends on the filter
length, and obviously heavily on the spacing and shift values. Another
important point is how you define the right-hand evaluation points of the
inversion ($r_\mr{def}$). Evaluating corresponding transform pairs separately
or jointly also leads to different filter coefficients ($J_0$, $J_1$, or $J_0$
\& $J_1$, or equally sine, cosine, or sine \& cosine); also if the real or the
imaginary part is used when complex transform pairs or \texttt{empymod} is
used; and if it is inverted for the minimum amplitude or for the maximum $r$.

@@

Figure~\ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his figure~5. The model consists
of a $0.3125\,\Omega\,$m sea layer of infinite thickness above a subsurface
$1\,\Omega\,$m half-space. The signal of a $1\,$Hz frequency $x$-directed
electric source 50\,m above the interface is measured at an $x$-directed
electric receiver at the interface.
%
\plot*{kongf5}{width=.9\textwidth}{(a) Results of different filters and QWE for
  the model of figure~5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of $10^{-12}$ and $10^{-30}$,
respectively. The relative error is shown in (b), where the QWE result was
taken as \emph{truth}. For offsets greater than roughly 15.5\,km the relative
error becomes meaningless, as the QWE fails itself; this part is greyed out in
the figure.

Figure~\ref{fig:keyf5}(a) considers the canonical CSEM model as used in
\cite{GEO.12.Key} in his figure~5: a $0.303\,\Omega\,$m sea with depth of
2\,km, a background resistivity of $1\,\Omega\,$m, in which a $100\,\Omega\,$m
target layer of 100\,m thickness is embedded at 1000\,m below the seafloor.
The source is located at 1990\,m and receivers are positioned on the seafloor.
%
\plot*{keyf5}{width=.9\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. The new filter has generally relative errors which are several
  orders of magnitude lower than the other filters.}
%

@@

Another error test is shown in Figure~\ref{fig:skindepth1}. It shows in (a) the
amplitude of the analytical fullspace solution for a wide range of
resistivities and frequencies: $\rho=$ $10^{-10}\,\Omega\,$m to
$10^{10}\,\Omega\,$m, and $f =10^{-10}$\,Hz to $10^{14}$\,Hz. In this case, the
horizontal offset is set to the skin depth, and the vertical source-receiver
separation to a tenth of it, where the skin depth is given by \citep[e.g., ][
equation~1.49]{B.SEG.88.Ward}
%
\begin{equation}
  z_\mr{skin} = \left\{\frac{\omega^2 \varepsilon\mu}{2}
    \left[ \sqrt{1 + \frac{1}{\left(\omega\varepsilon\rho\right)^2}} -1\right]
  \right\}^{-1/2} \, . \\
  \label{eq:skin}
\end{equation}
%
The grey contour lines in the figure show the magnitude of the offset: the line
with 0 is where the offset is 1\,m, the line with 3 where the offset is 1\,km,
etc. The contour lines of the offset have a knee, after which they become
almost independent of frequency. This is where the quasi static approximation
breaks down, and wave phenomena become dominant. The red line indicates where
$\omega\varepsilon\rho = 1$. Subplot (b) shows the relative error of the DLF
using the filter \emph{Wer201} compared to the analytical solution. It shows
nicely that the filter can be used for all cases that are in the quasi static
region, but not outside.
%
\plot*{skindepth1}{width=.8\textwidth}{(a) Amplitude of the analytical
  fullspace solution for a wide range of resistivities and frequencies. The
  contour lines show the power of the offset used for the calculation, where
  the offset is set to the skin depth. (b) Relative error of the DLF compared
  with the analytical solution. The filter \emph{Wer201} works well in the
  diffusive limit, but fails when wave phenomena become dominant. The red
 line shows where $\omega\varepsilon\rho = 1$.}
%

@@

Figure~\ref{fig:GPR} shows the GPR example as calculated in
\cite{GEO.15.Hunziker} and \cite{GEO.17.Werthmuller}; the model parameters are
given in subplot (a). The filter used for this example for the Hankel transform
is a 2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for
the inversion and the check of quality. The whole space solution used
resistivity $200\,\Omega\,$m, relative permittivity $\varepsilon_\mr{r}=10$,
and permeability $\mu_\mr{r}=1$. The source and receiver vertical separation is
1\,m. For the Fourier transform a 4096\,pt FFT was used with regularly spaced
frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up to 2048\,MHz. The
frequency result is multiplied with a Ricker wavelet with a center frequency of
250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is applied. Subplot (b)
shows the result when using the adaptive quadrature (QUAD) option of
\texttt{empymod}, which uses the QAGSE routine from the Fortran QUADPACK
library. Subplot (c) shows the result using DLF with the new 2001\,pt digital
filter. The calculation with DLF took under 9 minutes and is therefore roughly
80 times faster than the calculation with QUAD, which took roughly 11 hours and
27 minutes. Calculating the same model with QWE took 7 hours and 20 minutes.
However, QWE uses in roughly 1/3 of the calculation internally QUAD in this
example, see \cite{GEO.17.Werthmuller}. Note that DLF was run in parallel using
4 threads at once, taking effectively only 2 minutes and 10 seconds to
calculate.
% EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
% minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.
%
\plot*{GPR}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) the adaptive quadrature (QUAD) or with (c) a
  2001\,pt digital linear filter (b). QUAD took about 11\,h 27\,min to
  calculate, whereas DLF took 2\,min 10\,s in parallel on 4 nodes, hence under
  9\,min in total.}
%

>>>>

%%fakesection ===    ABSTRACT    ===
\begin{abstract} % 1-2 sentence(s) each
%
% (1) PRINCIPAL OBJECTIVES AND SCOPE OF THE WORK
  The open-source code \texttt{fdesign} makes it possible to design digital
  linear filters for the Hankel and Fourier transforms used in potential-,
  diffusive-, and wavefield modeling. Digital filters can be derived for any
  electromagnetic method, such as methods in the diffusive limits (direct
  current, controlled-source electromagnetics) as well as methods using higher
  frequency content (ground-penetrating radar, acoustic and elastic wavefields).
%
% (2) METHODOLOGY
  The direct matrix inversion method is used for the derivation of the filter
  values, and a brute-force inversion is carried out over the defined spacing
  and shifting values of the filter basis. Included or user-provided
  theoretical transform pairs are used for the inversion. Alternatively, one
  can provide layered subsurface models that will be computed with a precise
  quadrature method using the electromagnetic modeler \texttt{empymod} to
  generate numerical transform pairs.
%
% (3) RESULTS
  The comparison of the presented 201\,pt filter with previously presented
  filters shows that it performs better for some standard controlled-source
  electromagnetic cases. The derivation of a longer 2001\,pt filter for a
  ground-penetrating radar example with 250\,MHz center frequency proves that
  the filter method works also for wave phenomena, not only for diffusive
  electromagnetic fields.
%
% (4) CONCLUSIONS
  The presented algorithm provides a tool to create problem specific digital
  filters. Such purpose-built filters can be made shorter and can speed up
  consecutive  potential-, diffusive-, and wavefield inversions.
%
\end{abstract}

@@

\subsection{Minimization criteria}
Figure~\ref{fig:InvCrit} shows the differences between the different
minimization approaches; in (a) for a conventional, fast decaying transform
pair; and in (b) for a more complex, high frequency layered model. The circles
show the minimum amplitude used in previous approaches. This criterion is not
ideal, as it is subject to some random fluctuations and depends on the choice
of $r$. The squares show the minimum amplitude given a certain acceptable
error, and the diamonds show the maximum $r$ given a certain acceptable error
(the inversion is a minimization process, it therefore minimizes $1/r$, not
$r$). In simple cases the minimum amplitude given an acceptable error and the
maximum $r$ given an acceptable error will yield the same result. However, in
complex cases the maximum $r$ is more consistent and therefore a much better
criterion.
%
\plot*{InvCrit}{width=.9\textwidth}{(a) A regular RHS curve of a transform pair
  with a purely decaying function. Minimum amplitude or maximum $r$ yield the
  same result in this scenario. Using a relative error criteria is more stable
  than just the absolute minimum amplitude. (b) A 1D model for $f=200\,$MHz.
  Here, the maximum $r$ provides a better criteria for the inversion.}
%

@@

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often  a good filter is found by sheer luck of
choosing the right starting parameters. Each of the input parameters has an
effect on the outcome. It also depends heavily on the transform pairs
$f_\mr{I;C}$; functions that decay rapidly are generally better, as noted by
earlier authors \citep[e.g.][]{USGS.75.Anderson}. It also depends on the filter
length, and obviously heavily on the spacing and shift values. Another
important point is how you define the right-hand evaluation points of the
inversion ($r_\mr{def}$). Evaluating corresponding transform pairs separately
or jointly also leads to different filter coefficients ($J_0$, $J_1$, or $J_0$
\& $J_1$, or equally sine, cosine, or sine \& cosine); also if the real or the
imaginary part is used when complex transform pairs or \texttt{empymod} is
used; and if it is inverted for the minimum amplitude or for the maximum $r$.

@@

Figure~\ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his Figure~5. The model consists
of a $0.3125\,\Omega\,$m sea layer of infinite thickness above a subsurface
$1\,\Omega\,$m half-space. The signal of a $1\,$Hz frequency $x$-directed
electric source 50\,m above the interface is measured at an $x$-directed
electric receiver at the interface.
%
\plot*{kongf5}{width=.9\textwidth}{(a) Results of different filters and QWE for
  the model of Figure~5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of $10^{-12}$ and $10^{-30}$,
respectively. The relative error is shown in (b), where the QWE result was
taken as \emph{truth}. For offsets greater than roughly 15.5\,km the relative
error becomes meaningless, as the QWE fails itself; this part is greyed out in
the figure.

Figure~\ref{fig:keyf5}(a) considers the canonical CSEM model as used in
\cite{GEO.12.Key} in his Figure~5: a $0.303\,\Omega\,$m sea with depth of
2\,km, a background resistivity of $1\,\Omega\,$m, in which a $100\,\Omega\,$m
target layer of 100\,m thickness is embedded at 1000\,m below the seafloor.
The source is located at 1990\,m and receivers are positioned on the seafloor.
%
\plot*{keyf5}{width=.9\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. The new filter has generally relative errors which are several
  orders of magnitude lower than the other filters.}
%

@@

Another error test is shown in Figure~\ref{fig:skindepth1}. It shows in (a) the
amplitude of the analytical fullspace solution for a wide range of
resistivities and frequencies: $\rho=$ $10^{-10}\,\Omega\,$m to
$10^{10}\,\Omega\,$m, and $f =10^{-10}$\,Hz to $10^{14}$\,Hz. In this case, the
horizontal offset is set to the skin depth, and the vertical source-receiver
separation to a tenth of it, where the skin depth is given by \citep[e.g., ][
equation~1.49]{B.SEG.88.Ward}
%
\begin{equation}
  z_\mr{skin} = \left\{\frac{\omega^2 \varepsilon\mu}{2}
    \left[ \sqrt{1 + \frac{1}{\left(\omega\varepsilon\rho\right)^2}} -1\right]
  \right\}^{-1/2} \, , \\
  \label{eq:skin}
\end{equation}
%
where we used in the example $10\,\varepsilon_0$ for the electric permittivity
and $\mu_0$ for the magnetic permeability; $\omega = 2\pi f$. The grey contour
lines in the figure show the magnitude of the offset: the line with 0 is where
the offset is 1\,m, the line with 3 where the offset is 1\,km, etc. The contour
lines of the offset have a knee, after which they become almost independent of
frequency. This is where the quasi static approximation breaks down, and wave
phenomena become dominant. The red line indicates where $\omega\varepsilon\rho
= 1$. Subplot (b) shows the relative error of the DLF using the filter
\emph{Wer201} compared to the analytical solution. It shows nicely that the
filter can be used for all cases that are in the quasi static region, but not
outside.
%
\plot*{skindepth1}{width=.8\textwidth}{(a) Amplitude of the analytical
  fullspace solution for a wide range of resistivities and frequencies. The
  contour lines show the power of the offset used for the calculation, where
  the offset is set to the skin depth. (b) Relative error of the DLF compared
  with the analytical solution. The filter \emph{Wer201} works well in the
  diffusive limit, but fails when wave phenomena become dominant. The red
 line shows where $\omega\varepsilon\rho = 1$.}
%

@@

Figure~\ref{fig:GPR} shows the GPR example as calculated in
\cite{GEO.15.Hunziker} and \cite{GEO.17.Werthmuller}; the model parameters are
given in subplot (a). The filter used for this example for the Hankel transform
is a 2001\,pt filter, derived with the fullspace solution with $f=500\,$MHz for
the inversion and the check of quality. The whole space solution used
resistivity $200\,\Omega\,$m, relative permittivity $\varepsilon_\mr{r}=10$,
and permeability $\mu_\mr{r}=1$. Using Equation~\ref{eq:skin} yields a skin
depth of roughly 3.6\,m for these parameters. The source and receiver vertical
separation is 1\,m. For the Fourier transform a 4096\,pt FFT was used with
regularly spaced frequencies from 0.5\,MHz to 850\,MHz and then zero-padded up
to 2048\,MHz. The frequency result is multiplied with a Ricker wavelet with a
center frequency of 250\,MHz, and a gain function ($1 + |t^3|, t$ in ns) is
applied. Subplot (b) shows the result when using the adaptive quadrature (QUAD)
option of \texttt{empymod}, which uses the QAGSE routine from the Fortran
QUADPACK library. Subplot (c) shows the result using DLF with the new 2001\,pt
digital filter. The calculation with DLF took under 9 minutes and is therefore
roughly 80 times faster than the calculation with QUAD, which took roughly 11
hours and 27 minutes. Calculating the same model with QWE took 7 hours and 20
minutes. However, QWE uses in roughly 1/3 of the calculation internally QUAD in
this example, see \cite{GEO.17.Werthmuller}. Note that DLF was run in parallel
using 4 threads at once, taking effectively only 2 minutes and 10 seconds to
calculate.
% EMmod \citep{GEO.15.Hunziker} took for the same model about 18 hours 55
% minutes, using a 61\,pt Gauss-Kronrod integration routine.
The lagged convolution version of DLF and the splined version of QWE were used
in this comparison.
%
\plot*{GPR}{width=\textwidth}{GPR example for the model given in (a) using
  \texttt{empymod} with (b) the adaptive quadrature (QUAD) or with (c) a
  2001\,pt digital linear filter (b). QUAD took about 11\,h 27\,min to
  calculate, whereas DLF took 2\,min 10\,s in parallel on 4 nodes, hence under
  9\,min in total.}
%
